Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer

### 2020-04-14 manuscript/manuscript.tex
updated figures and addressed most of chloé's comments

\begin{abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. Indeed, it allows to jointly consider the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study the utility of six methods that identify subnetworks with higher association scores to a phenotype to discover new biomarkers for breast cancer susceptibility. We interrogate a familial breast cancer genome-wide association study (GWAS) focused on \emph{BRCA1/2} negative French women. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical stringency for biological meaningfulness, most network methods give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. Interestingly, a combination of solution subnetworks provided a concise subnetwork of 93 genes, enriched in known breast cancer susceptibility genes (\emph{BABAM1}, \emph{BLM}, \emph{CASP8}, \emph{FGFR2}, and \emph{TOX3}, Fisher's exact test P-value~=~$7.8 \times 10^{-5}$) and occupying more central positions in the network than average. Additionally, it shows a general alteration of the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Finally, we find a significantly large overlap between the genes in the consensus network and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely lack of statistical power and of interpretable solutions.
\cazcom{}{Add a sentence to the effect that the computational tools are available on github for ease of application to other data sets?}
\end{abstract}

@@

In human health, genome-wide association studies (GWAS) aim at quantifying how single-nucleotide polymorphisms (SNPs) predispose to complex diseases, like diabetes or some forms of cancer \cite{bush_chapter_2012}. To that end, in a typical GWAS thousands of unrelated samples are genotyped: the cases, suffering the disease of interest, and the controls from the general population. Then, a per-SNP statistical association test is conducted (e.g. \cazcom{\caz{}{based on }logistic regression}{You're doing a chi2 allelic test later on, so maybe use this here rather than talking about logistic regression?}). Those SNPs with a P-value lower than a conservative Bonferroni threshold are candidates to further studies in an independent cohort. Once the risk SNPs have been discovered, they can be used for risk assessment, and to deepen our understanding of the disease.

@@

Network-based biomarker discovery methods exploit this relatedness to identify disease genes on GWAS data \cite{azencott_network-guided_2016}. In essence, each SNP has a score of association with the disease, given by the experiment, and functionally biological relationships, given by a network built on prior knowledge. Then, the problem becomes finding a functionally-related set of highly-scoring genes. Different solutions have been proposed to this problem, often stemming from divergent mathematical frameworks and considerations of what the optimal solution looks like. Some methods strongly constrain the problem to certain kinds of subnetworks. Such is the extreme case of LEAN \cite{gwinner_network-based_2016}, which focuses on ``star'' subnetworks, i.e. instances were both a gene and its direct interactors are associated with the disease. Other algorithms, like dmGWAS \cite{jia_dmgwas:_2011} and heinz \cite{dittrich_identifying_2008}, focus on larger subnetworks interconnecting genes with high association scores. However, they differ in their tolerance to the inclusion of low-scoring nodes, and the possible number of \cazcom{\caz{disconnected subnetworks}{connected components}}{Are you worried people don't understand ``connected components''? If so, we should define it and use it.} in the solution. Lastly, other methods also consider the topology of the network, favoring groups of nodes that are not only high-scoring, but also densely interconnected; such is the case of HotNet2 \cite{leiserson_pan-cancer_2015}, SConES \cite{azencott_efficient_2013}, and SigMod \cite{liu_sigmod:_2017}.

@@

\caz{On top of}{In addition to} a weight, nodes have other properties, provided by the topology of the graph. In this article we focus on two\caz{}{of those}: degree centrality, and betweenness centrality. The degree centrality, or degree, is the number of edges that a node has. The betweenness centrality, or betweenness, is the number of times a node participates in the shortest paths between two other nodes.

In addition, we use several matrices that describe different properties of a graph. The described matrices are square, and have as many rows and columns as nodes are in the network. The element $(i,j)$ represents a  selected relationship between $v_i$ and $v_j$. The \emph{adjacency matrix} $W_G$ contains a 1 when the corresponding nodes are connected, and 0 otherwise; the diagonal is zero. The \emph{degree matrix} $D_G$ is a diagonal matrix which contains the degree of the different nodes. Lastly, the \emph{Laplacian matrix} $L_G$ is defined as $L_G = D_G - W_G$.

@@

\paragraph{Gene network}
Out of the six methods tested, \cazcom{five use a gene-gene interaction network}{SConES GI does too...} (Section~\ref{methods:methods}). Although their respective statistical frameworks are compatible with any type of network (protein interactions, gene coexpression, regulatory, etc.), we used protein-protein interaction networks (PPIN), as they are interpretable, well characterized, and most of the methods were designed to scale appropriately to them. We built our PPIN from both binary and co-complex interactions stored in the HINT database (release April 2019) \cite{das_hint:_2012}. Unless otherwise specified, we used only interactions coming from high-throughput experiments, leaving out targeted studies that might bias the topology of the network. Out of the 146\,722 interactions from high-throughput experiments that HINT stores, we were able to map 142\,541 to a pair of gene symbols. The scoring function for the nodes changed from method to method (Section~\ref{methods:methods}).

Additionally, we compared the results of the aforementioned PPIN with those obtained on another PPIN built using interactions coming from both high-throughput and targeted studies. In that case, out of the 179\,332 interactions in HINT, 173\,797 mapped to a pair of gene symbols.

\cazcom{}{How many genes are involved in those interactions? In the ``SNP and gene association'' section, you describe 164\,037 SNPs mapping to 32\,767 gene symbols. How many of these genes and SNPs are involved in the aforementioned 142\,541 (resp. 173\,797) interactions? And what happens to the genes to which you have mapped SNPs but that are not involved in any HINT association: are they discarded, or do they remain as isolated nodes?}

\paragraph{SNP networks}
SConES \cite{azencott_efficient_2013} is the only studied methods designed to handle SNP networks. As in gene networks, two SNPs are linked in a SNP network when there is evidence of shared functionality between two SNPs. \citet{azencott_efficient_2013} proposed three ways of building these networks: connecting the SNPs consecutive in the genomic sequence (``GS network''); interconnecting all the SNPs mapped to the same gene, on top of GS (``GM network''); and interconnecting all SNPs mapped to two genes for which a protein-protein interaction exists, on top of GM (``GI network''). We focused on the GI network, as it fits the scope of this work better, using the PPIN described above. However, at different stages of this work we also used GS and GM for comparison. For the GM network, we used the mapping described in Section~\ref{methods:node_score}. In all three the node scores are the association scores of the individual SNPs with the phenotype (1~d.f.~\(\chi\)\textsuperscript{2}).

\cazcom{}{What are the sizes of the resulting GS, GM and GI networks? 164\,037 SNPs/nodes (or fewer if some aren't used on the gene-gene networks and you've removed them entirely from the analysis)? How many edges in the different networks? If you're worried about space, the sizes of the gene-gene and SNP-SNP networks from the previous paragraph and this one could be given in a table in the supplementaries.}

\subsubsection{Studied methods}
\label{methods:methods}
\cazcom{}{Maybe we could change ``studied'' in this section (title, text, tables...) to something that makes it sound a bit less like a school assignment, and also less like we decided to study those but could have studied others? Something like ``Network-based methods'' would have a bit more agency to it? Sorry if I'm nitpicking.}

\begin{table}[htbp]
  \caption{\label{tab:method_comparison} Summary of the differences between the studied algorithms.}
  \centering
  \begin{threeparttable}
    \begin{tabular}{l|llllllll}
      Method & Field & Nodes & Exhaustive & Solution & Comp. & Input & Scoring & Ref.\\
      \hline
      dmGWAS & GWAS & Genes & No & - & 1 & Summary & -log\textsubscript{10}(P) & \cite{jia_dmgwas:_2011}\\
      heinz & Omics & Genes & Yes & - & 1 & Summary & BUM & \cite{dittrich_identifying_2008}\\
      HotNet2 & Omics & Genes & Yes & Module & \(\ge\) 1 & Summary & Local FDR & \cite{leiserson_pan-cancer_2015}\\
      LEAN & Omics & Genes & Yes & Star & \(\ge\) 1 & Summary & -log\textsubscript{10}(P) & \cite{gwinner_network-based_2016}\\
      SConES & GWAS & SNPs & Yes & Module & \(\ge\) 1 & Genotypes & 1 d.f. \(\chi\)\textsuperscript{2} & \cite{azencott_efficient_2013}\\
      SigMod & GWAS & Genes & Yes & Module & 1 & Summary & -log\textsubscript{10}(P) & \cite{liu_sigmod:_2017}\\
    \end{tabular}
    \begin{tablenotes}
      \footnotesize{
        \item \textbf{\emph{Field}}: field in which the algorithm was developed. \textbf{\emph{Nodes}}: the type of network, either gene (protein-protein interaction network usually) or a SNP network. \textbf{\emph{Exhaustive}}: whether all the possible solutions given the selected hyperparameters are explored. \textbf{\emph{Solution}}: additional properties are enforced on the solution subnetwork, other than being dense in high scores and connected. \textbf{\emph{Comp.}}: number of connected \caz{subnetworks}{components} in the solution. \textbf{\emph{Input}}: genotype data or GWAS summary statistics. \textbf{\emph{Scoring}}: how SNP/gene P-values are transformed into node scores. \textbf{\emph{Ref.}}: original publication featuring the algorithm.
      }
    \end{tablenotes}
  \end{threeparttable}
\end{table}

@@

\begin{enumerate}
\item Select a seed node \caz{}{$i$ and form $S_i=\{i\}$}.
\item Compute Stouffer's Z-score Z\textsubscript{m} for the current subgraph \caz{$S$}{$S_i$} as

@@

where \emph{k} is the number of genes in $S_i$; z\textsubscript{j} is the Z score of gene $j$, computed as \(\phi\)\textsuperscript{-1}(1 - P-value\textsubscript{j}); and \(\phi\)\textsuperscript{-1} is the inverse normal distribution function.
\item Identify neighboring nodes of \caz{$S$}{$S_i$},  i.e. nodes at distance \(\le\) d.
\item Add the neighboring nodes whose inclusion increases the Z\textsubscript{m+1} more than a threshold Z\textsubscript{m}~\texttimes{}~(1 + r).
\item Repeat 2-4 until no further enlargement is possible.
\caz{}{\item Add $S_i$ to the list of subnetworks to return.}
\end{enumerate}
\cazcom{}{Missing here: how to select seed nodes. Is every node a seed node? Every node that has not yet been included in a subgraph?}

Lastly, a subnetwork's Z-score is normalized as

@@

where Z\textsubscript{m}(\(\pi\)) represents a vector containing 100\,000 random subsets of the same number of genes.

We used the implementation of dmGWAS in the dmGWAS 3.0 R package \cite{dmgwas}. We used the suggested hyperparameters $d = 2$ and $r = 0.1$. We used the function \emph{simpleChoose} to select the solution subnetwork, which aggregates the top 1\% modules into the solution subnetwork.
\end{description}

@@

were $p(v)~=~w(v) - w'$ is the \emph{profit} of adding a node, \caz{$c(e)~=~w'$}{$c(e)=-w'$} is the \emph{cost} of adding an edge, and $w' = min_{v \in V_{G}} w(v)$ is the smallest node weight of $G$. All three are positive quantities. Heinz implements the algorithm from \citet{ljubic_algorithmic_2006} which, in practice is often fast and optimal, although neither is guaranteed. We used BioNet's implementation of heinz \cite{beisser_bionet:_2010,heinz}.

@@

% \begin{equation}
% \label{eq:scones}
% \underset{S \subseteq G}{\arg \max } \underbrace{\sum_{v \in V_S} w(v)}_{\text { association }} + \underbrace{\lambda \sum_{v \in V_S} \sum_{u \not\in V_S} L_{vu} }_{\text { connectivity }}-\underbrace{\eta \lvert V_S \rvert }_{\text { sparsity }}
% \end{equation}

\cazcom{}{Your equation for SConES was wrong.
Use either of the following (I'd use the one that does not involve the Laplacian, then you do not need to introduce the Laplacian in the Mathematical notations.
One way to look at it is to start from the bottom one, where the connectivity term increases with each edge that connects a selected node to a non-selected node, and as you're maximizing the whole thing, you need a minus in top of it. Then 
\[
\sum_{u \in V_S} \sum_{v \not\in V_S} W_{uv} = \sum_{u \in V_S} \left(\sum_{v \in V} W_{uv} - \sum_{v \in V_S} W_{uv} \right) = \sum_{u \in V_S} \left(D_{uu} - \sum_{v \in V_S} W_{uv} \right) 
= \sum_{u \in V_S} \sum_{v \in V_S} \left(D_{uv} - W_{uv} \right) \]
The first equality is true because the set of all nodes is the union of the set of nodes of $S$ and the set of nodes not in $S$; the second by definition of the degree; and the third because the degree matrix is diagonal and $D_{uv}=0$ for $u \neq v$. Finally $L_{uv} = D_{uv} - W_{uv}$.
}

\begin{equation}
\label{eq:scones}
\underset{S \subseteq G}{\arg \max } \underbrace{\sum_{v \in V_S} w(v)}_{\text { association }} {\color{red}-} \underbrace{\lambda \sum_{v \in V_S} \sum_{u {\color{red} \in} V_S} L_{vu} }_{\text { connectivity }}-\underbrace{\eta \lvert V_S \rvert }_{\text { sparsity }}
\end{equation}

or

\begin{equation}
\label{eq:scones}
\underset{S \subseteq G}{\arg \max } \underbrace{\sum_{v \in V_S} w(v)}_{\text { association }} {\color{red}-} \underbrace{\lambda \sum_{v \in V_S} \sum_{u \not\in V_S} {\color{red}W}_{vu} }_{\text { connectivity }}-\underbrace{\eta \lvert V_S \rvert }_{\text { sparsity }}
\end{equation}

@@

SigMod presents three important differences with SConES. First it is designed for gene-gene networks. Second, it favors subnetworks containing many edges. SConES, instead, penalizes connections between the selected and unselected nodes. Third, it returns a single connected subnetwork, which it achieves by exploring a grid of hyperparameters and processing their respective solutions. Specifically, for the range of \(\lambda\)~=~\(\lambda\)\textsubscript{min}, \dots{}, \(\lambda\)\textsubscript{max} for the same \(\eta\), it prioritizes the solution with the largest change in size from \(\lambda\)\textsubscript{n} to \(\lambda\)\textsubscript{n+1}. Such a large change implies that the network is densely interconnected. This results in one candidate solution for each \(\eta\), which are processed by removing any node not connected to any other. A score is assigned to each candidate solution by summing their node scores and normalizing by size. The candidate solution with the highest standardized score is the chosen solution. SigMod is implemented in an R package \cite{sigmod}.
\end{description}

@@

We searched for pathways enriched in the gene subnetworks produced by the above methods. We conducted an hypergeometric test on pathways from Reactome \cite{Jassal2019} using the R package ReactomePA \cite{Yu2016}. The universe of genes included any gene that we could map to a SNP in the iCOGS array (Section \ref{methods:node_score}). Pathways with an adjusted P-value \textless~0.05 were deemed significant.

@@

We compared the runtime, the number of selected genes/SNPs, and the stability (sensitivity of the result to small changes in the input, here, using different subsets of the data) of the different network methods. The stability was measured using Pearson correlation between different runs as suggested by \citet{nogueira_measuring_2016}.

@@

A desirable solution offers good predictive power on unseen \emph{test} samples. We evaluated the predicting power of the SNPs selected by the different methods through the performance of an L1-penalized logistic regression classifier trained exclusively on those SNPs to predict the outcome (case/control). The L1 penalty helps to account for LD by reducing the number of SNPs included in the model (size of the active set), while improving the generalization of the classifier. This penalty was set by cross-validation. We applied each network method to each \emph{train} set, and trained the classifier on it as well using only on the selected SNPs. When the method retrieved a list of genes (all of them except SConES), we considered as selected all the SNPs mapped to any of those genes. Then we evaluated the sensitivity and the specificity on the \emph{test set}. The active set gave an estimate of a plausible, more sparse solution with a comparable predictive power to the original solution. To obtain a baseline, we also trained the classifier on all the SNPs.

@@

Closer examination reveals two regions (3p24 and 8q24) having low, albeit not genome-wide significant, P-values. Both of them have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression on the whole dataset (Section~\ref{methods:classifier}), \cazcom{a machine learning algorithm that searches a small subset of SNPs which provide good classification accuracy}{describe it in the methods rather than here? How was the regularization parameter chosen?}. It selected 100 SNPs, both from all aforementioned regions and from other regions (Supplementary Figure~\ref{sfig:snp_gene_manhattan}C). Yet, it is unclear why those SNPs were selected, as emphasized by the high P-value of some of them, which further complicates the biological interpretation. Moreover, and in opposition to what the omnigenic model predicts, the genes to which this SNPs map to (Section~\ref{methods:node_score}) are not interconnected in the PPIN (Section~\ref{methods:networks}). \caz{}{In addition, the classification performance of the method is very low, and L1-penalized methods select only one of several correlated variables and are prone to instability, which further complicates interpretation.} This motivates exploring network methods, which trade statistical significance for biological relevance to find susceptibility subnetworks. In fact, such methods provided comparable classification performance to L1-penalized logistic regression \cazcom{(Figure~\ref{fig:benchmark}B)}{Is it B? I don't seel L1-penalized logistic regression on this figure; also with these types of performance I'm not sure it's worth making that point. Maybe frame it as ``do no worse'' than ``have comparable performance''...}, while providing more interpretable solutions.

\subsection{Network methods successfully identify genes associated with breast cancer}
\label{results:separate_networks}
\begin{table}[htbp]
  \centering
  \begin{threeparttable}
\caption{\label{tab:gene_solutions}
Summary statistics on the \caz{results}{outputs/solutions} of multiple network methods \caz{on the gene-gene interaction network}{}. The first row contains the summary statistics on the whole network.}
\centering
\begin{tabular}{lrrrrr}
Network & \# genes & \# edges & \mean{Betweenness} & \median{P}\textsubscript{gene} & \(\rho\)\textsubscript{consensus}\\
\hline
HINT HT & 13\,619 & 142\,541 & 16\,706 & 0.46 & 0.066\\
\hline
Consensus & 55 & 117 & 74\,062 & 0.0051 & 1\\
dmGWAS & 194 & 450 & 49\,115 & 0.19 & 0.41\\
heinz & 4 & 3 & 113\,633 & 0.0012 & 0.21\\
HotNet2 & 440 & 374 & 7\,739 & 0.048 & 0.31\\
LEAN & 0 & 0 & - & - & -\\
SConES GI & 0 (1) & 0 & - & - & -\\
SigMod & 142 & 249 & 92\,603 & 0.0083 & 0.73\\
\end{tabular}
\begin{tablenotes}
  \footnotesize{
    \item \textbf{\emph{\# genes}}: number of genes selected out of those that are part of the PPIN; for SConES GI the total number of genes, including RNA genes, was added in parentheses. \textbf{\mean{Betweenness}}: mean betweenness of the selected genes in the PPIN. \textbf{\median{P}\textsubscript{gene}}: median P-value of the selected genes. \textbf{\(\rho\)\textsubscript{consensus}}: Pearson correlation between the subnetwork and the consensus network.
  }
\end{tablenotes}
\end{threeparttable}
\end{table}

>>>>

\begin{abstract}
Systems biology provides a comprehensive approach to biomarker discovery and biological hypothesis building. Indeed, it allows to jointly consider the statistical association between gene variation and a phenotype, and the biological context of each gene, represented as a network. In this work, we study the utility of six methods that identify subnetworks with higher association scores to a phenotype to discover new biomarkers for breast cancer susceptibility. We interrogate a familial breast cancer genome-wide association study (GWAS) focused on \emph{BRCA1/2} negative French women. We perform an in-depth benchmarking of the methods with regards to size of the solution subnetwork, their utility as biomarkers, and the stability and the runtime of the methods. By trading statistical stringency for biological meaningfulness, most network methods give more compelling results than standard SNP- and gene-level analyses, recovering causal subnetworks tightly related to cancer susceptibility. Interestingly, a combination of solution subnetworks provided a concise subnetwork of 93 genes, enriched in known breast cancer susceptibility genes (\emph{BABAM1}, \emph{BLM}, \emph{CASP8}, \emph{FGFR2}, and \emph{TOX3}, Fisher's exact test P-value~=~$7.8 \times 10^{-5}$) and occupying more central positions in the network than average. Additionally, it shows a general alteration of the neighborhood of \emph{COPS5}, a gene related to multiple hallmarks of cancer. Finally, we find a significantly large overlap between the genes in the consensus network and the genes significantly associated in the largest GWAS on susceptibility to breast cancer. In conclusion, this article shows the pertinence of network-based analyses to tackle known issues with GWAS, namely lack of statistical power and of interpretable solutions. Project-agnostic implementations of each of the network methods are available at \url{https://github.com/hclimente/gwas-tools} to facilitate their application to other GWAS datasets.
\end{abstract}

@@

In human health, genome-wide association studies (GWAS) aim at quantifying how single-nucleotide polymorphisms (SNPs) predispose to complex diseases, like diabetes or some forms of cancer \cite{bush_chapter_2012}. To that end, in a typical GWAS thousands of unrelated samples are genotyped: the cases, suffering the disease of interest, and the controls from the general population. Then, a per-SNP statistical association test is conducted (e.g. based on logistic regression). Those SNPs with a P-value lower than a conservative Bonferroni threshold are candidates to further studies in an independent cohort. Once the risk SNPs have been discovered, they can be used for risk assessment, and to deepen our understanding of the disease.

@@

Network-based biomarker discovery methods exploit this relatedness to identify disease genes on GWAS data \cite{azencott_network-guided_2016}. In essence, each SNP has a score of association with the disease, given by the experiment, and functionally biological relationships, given by a network built on prior knowledge. Then, the problem becomes finding a functionally-related set of highly-scoring genes. Different solutions have been proposed to this problem, often stemming from divergent mathematical frameworks and considerations of what the optimal solution looks like. Some methods strongly constrain the problem to certain kinds of subnetworks. Such is the extreme case of LEAN \cite{gwinner_network-based_2016}, which focuses on ``star'' subnetworks, i.e. instances were both a gene and its direct interactors are associated with the disease. Other algorithms, like dmGWAS \cite{jia_dmgwas:_2011} and heinz \cite{dittrich_identifying_2008}, focus on larger subnetworks interconnecting genes with high association scores. However, they differ in their tolerance to the inclusion of low-scoring nodes, and the topology of the solution. Lastly, other methods also consider the topology of the network, favoring groups of nodes that are not only high-scoring, but also densely interconnected; such is the case of HotNet2 \cite{leiserson_pan-cancer_2015}, SConES \cite{azencott_efficient_2013}, and SigMod \cite{liu_sigmod:_2017}.

@@

In addition to a weight, nodes have other properties, provided by the topology of the graph. In this article we focus on two of those: degree centrality, and betweenness centrality. The degree centrality, or degree, is the number of edges that a node has. The betweenness centrality, or betweenness, is the number of times a node participates in the shortest paths between two other nodes.

In addition, we use several matrices that describe different properties of a graph. The described matrices are square, and have as many rows and columns as nodes are in the network. The element $(i,j)$ represents a  selected relationship between $v_i$ and $v_j$. The \emph{adjacency matrix} $W_G$ contains a 1 when the corresponding nodes are connected, and 0 otherwise; the diagonal is zero. The \emph{degree matrix} $D_G$ is a diagonal matrix which contains the degree of the different nodes.

@@

\paragraph{Gene network}
The statistical frameworks of the different network methods are compatible with any type of network (protein interactions, gene coexpression, regulatory, etc.). Yet, we used protein-protein interaction networks (PPIN) for all of them except SConES, as they are interpretable, well characterized, and they were designed to scale appropriately to them. We built our PPIN from both binary and co-complex interactions stored in the HINT database (release April 2019) \cite{das_hint:_2012}. Unless otherwise specified, we used only interactions coming from high-throughput experiments, leaving out targeted studies that might bias the topology of the network. Out of the 146\,722 interactions from high-throughput experiments that HINT stores, we were able to map 142\,541 to a pair of gene symbols, involving 13\,619 genes. 12\,880 of those mapped to a genotyped SNP after quality control, involving 127\,604 interactions. The scoring function for the nodes changed from method to method (Section~\ref{methods:methods}).

Additionally, we compared the results of the aforementioned PPIN with those obtained on another PPIN built using interactions coming from both high-throughput and targeted studies. In that case, out of the 179\,332 interactions in HINT, 173\,797 mapped to a pair of gene symbols. Out of those, 13\,735 mapped to a genotyped SNP after quality control, involving 156\,190 interactions.

\paragraph{SNP networks}
SConES \cite{azencott_efficient_2013} is the only network method designed to handle SNP networks. As in gene networks, two SNPs are linked in a SNP network when there is evidence of shared functionality between two SNPs. \citet{azencott_efficient_2013} proposed three ways of building these networks: connecting the SNPs consecutive in the genomic sequence (``GS network''); interconnecting all the SNPs mapped to the same gene, on top of GS (``GM network''); and interconnecting all SNPs mapped to two genes for which a protein-protein interaction exists, on top of GM (``GI network''). We focused on the GI network, as it fits the scope of this work better, using the PPIN described above. However, at different stages of this work we also used GS and GM for comparison. For the GM network, we used the mapping described in Section~\ref{methods:node_score}. In all three the node scores are the association scores of the individual SNPs with the phenotype (1~d.f.~\(\chi\)\textsuperscript{2}). The properties of these three subnetworks is available in Supplementary Table \ref{tab:snp_solutions}.

\subsubsection{Network methods}
\label{methods:methods}

\begin{table}[htbp]
  \caption{Summary of the differences between the network methods.}
  \label{tab:method_comparison}
  \centering
  \begin{threeparttable}
    \begin{tabular}{l|llllllll}
      Method & Field & Nodes & Exhaustive & Solution & Comp. & Input & Scoring & Ref.\\
      \hline
      dmGWAS & GWAS & Genes & No & - & 1 & Summary & -log\textsubscript{10}(P) & \cite{jia_dmgwas:_2011}\\
      heinz & Omics & Genes & Yes & - & 1 & Summary & BUM & \cite{dittrich_identifying_2008}\\
      HotNet2 & Omics & Genes & Yes & Module & \(\ge\) 1 & Summary & Local FDR & \cite{leiserson_pan-cancer_2015}\\
      LEAN & Omics & Genes & Yes & Star & \(\ge\) 1 & Summary & -log\textsubscript{10}(P) & \cite{gwinner_network-based_2016}\\
      SConES & GWAS & SNPs & Yes & Module & \(\ge\) 1 & Genotypes & 1 d.f. \(\chi\)\textsuperscript{2} & \cite{azencott_efficient_2013}\\
      SigMod & GWAS & Genes & Yes & Module & \(\ge\) 1 & Summary & $\Phi$\textsuperscript{-1}(1 - P) & \cite{liu_sigmod:_2017}\\
    \end{tabular}
    \begin{tablenotes}
      \footnotesize{
        \item \textbf{\emph{Field}}: field in which the algorithm was developed. \textbf{\emph{Nodes}}: the type of network, either gene (protein-protein interaction network usually) or a SNP network. \textbf{\emph{Exhaustive}}: whether all the possible solutions given the selected hyperparameters are explored. \textbf{\emph{Solution}}: additional properties are enforced on the solution subnetwork, other than being dense in high scores and connected. \textbf{\emph{Comp.}}: number of connected components in the solution. \textbf{\emph{Input}}: genotype data or GWAS summary statistics. \textbf{\emph{Scoring}}: how SNP/gene P-values are transformed into node scores. In the case of SigMod, $\Phi$\textsuperscript{-1} represents the inverse of the
        cumulative distribution function of the standard Normal distribution. \textbf{\emph{Ref.}}: original publication featuring the algorithm.
      }
    \end{tablenotes}
  \end{threeparttable}
\end{table}

@@

\begin{enumerate}
\item Select a seed node $i$ and form the subnetwork $S_i=\{i\}$.
\item Compute Stouffer's Z-score Z\textsubscript{m} for the current subgraph $S_i$ as

@@

where \emph{k} is the number of genes in $S_i$; z\textsubscript{j} is the Z score of gene $j$, computed as \(\phi\)\textsuperscript{-1}(1 - P-value\textsubscript{j}); and \(\phi\)\textsuperscript{-1} is the inverse normal distribution function.
\item Identify neighboring nodes of $S_i$,  i.e. nodes at distance \(\le\) d.
\item Add the neighboring nodes whose inclusion increases the Z\textsubscript{m+1} more than a threshold Z\textsubscript{m}~\texttimes{}~(1 + r).
\item Repeat 2-4 until no further enlargement is possible.
\item Add $S_i$ to the list of subnetworks to return. Its Z-score is normalized as

@@

where Z\textsubscript{m}(\(\pi\)) represents a vector containing 100\,000 random subsets of the same number of genes.
\end{enumerate}

DmGWAS carries out this process on every gene in the network. We used the implementation of dmGWAS in the dmGWAS 3.0 R package \cite{dmgwas}. We used the suggested hyperparameters $d = 2$ and $r = 0.1$. We used the function \emph{simpleChoose} to select the solution subnetwork, which aggregates the top 1\% modules into the solution subnetwork.
\end{description}

@@

were $p(v)~=~w(v) - w'$ is the \emph{profit} of adding a node, $c(e)~=~w'$ is the \emph{cost} of adding an edge, and $w' = min_{v \in V_{G}} w(v)$ is the smallest node weight of $G$. All three are positive quantities. Heinz implements the algorithm from \citet{ljubic_algorithmic_2006} which, in practice is often fast and optimal, although neither is guaranteed. We used BioNet's implementation of heinz \cite{beisser_bionet:_2010,heinz}.

@@

\begin{equation}
\label{eq:scones}
\underset{S \subseteq G}{\arg \max } \underbrace{\sum_{v \in V_S} w(v)}_{\text { association }} - \underbrace{\lambda \sum_{v \in V_S} \sum_{u \not\in V_S} {W}_{vu} }_{\text { connectivity }}-\underbrace{\eta \lvert V_S \rvert }_{\text { sparsity }}
\end{equation}

@@

SigMod presents three important differences with SConES. First it is designed for gene-gene networks. Second, it favors subnetworks containing many edges. SConES, instead, penalizes connections between the selected and unselected nodes. Third, it explores the grid of hyperparameters differently, and processes their respective solutions. Specifically, for the range of \(\lambda\)~=~\(\lambda\)\textsubscript{min}, \dots{}, \(\lambda\)\textsubscript{max} for the same \(\eta\), it prioritizes the solution with the largest change in size from \(\lambda\)\textsubscript{n} to \(\lambda\)\textsubscript{n+1}. Such a large change implies that the network is densely interconnected. This results in one candidate solution for each \(\eta\), which are processed by removing any node not connected to any other. A score is assigned to each candidate solution by summing their node scores and normalizing by size. The candidate solution with the highest standardized score is the chosen solution. SigMod is implemented in an R package \cite{sigmod}.
\end{description}

@@

We searched for pathways enriched in the gene subnetworks produced by the above methods. We conducted an hypergeometric test on pathways from Reactome \cite{Jassal2019} using the R package ReactomePA \cite{Yu2016}. The universe of genes included any gene that we could map to a SNP in the iCOGS array (Section \ref{methods:node_score}). Pathways with an adjusted P-value \textless~0.05 were deemed significant. As the significant pathways are often overlapping and redundant, the results were manually curated afterwards.

@@

We compared the runtime, the number of selected genes/SNPs, and the stability (sensitivity of the result to small changes in the input, here, using different subsets of the data) of the different network methods. The stability was quantified using the Pearson correlation to measure the overlap between different runs as suggested by \citet{nogueira_measuring_2016}.

@@

A desirable solution offers good predictive power on unseen \emph{test} samples. We evaluated the predicting power of the SNPs selected by the different methods through the performance of an L1-penalized logistic regression classifier, a machine learning algorithm that searches a small subset of SNPs which provide good classification accuracy. We trained the classifier exclusively on those selected SNPs to predict the outcome (case/control). The L1 penalty helps to account for LD by reducing the number of SNPs included in the model (\emph{active set}), while improving the generalization of the classifier. This penalty was set by cross-validation, choosing the value that minimized misclassification error. We applied each network method to each \emph{train} set, and trained the classifier on it as well using only on the selected SNPs. When the method retrieved a list of genes (all of them except SConES), we considered as selected all the SNPs mapped to any of those genes. Then we evaluated the sensitivity and the specificity on the \emph{test set}. The active set gave an estimate of a plausible, more sparse solution with a comparable predictive power to the original solution. To obtain a baseline, we also trained the classifier on all the SNPs.

@@

Closer examination reveals two regions (3p24 and 8q24) having low, albeit not genome-wide significant, P-values. Both of them have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression on the whole dataset (Section~\ref{methods:classifier}). It selected 100 SNPs, both from all aforementioned regions and from other regions (Supplementary Figure~\ref{sfig:snp_gene_manhattan}C). Yet, it is unclear why those SNPs were selected, as emphasized by the high P-value of some of them, which further complicates the biological interpretation. Moreover, and in opposition to what the omnigenic model predicts, the genes to which this SNPs map to (Section~\ref{methods:node_score}) are not interconnected in the PPIN (Section~\ref{methods:networks}). In addition, the classification performance of the method is very low, and L1-penalized methods select only one of several correlated variables and are prone to instability, which further complicates interpretation. This motivates exploring network methods, which trade statistical significance for biological relevance to find susceptibility subnetworks. In fact, such methods provided comparably (poor) classification performance to L1-penalized logistic regression (Figure~\ref{fig:benchmark}B), while providing more interpretable solutions.

\subsection{Network methods successfully identify genes associated with breast cancer}
\label{results:separate_networks}
\begin{table}[htbp]
  \centering
  \begin{threeparttable}
\caption{Summary statistics on the solutions of multiple network methods on the gene-gene interaction network. The first row contains the summary statistics on the whole network.}
\label{tab:gene_solutions}
\centering
\begin{tabular}{lrrrrr}
Network & \# genes & \# edges & \mean{Betweenness} & \median{P}\textsubscript{gene} & \(\rho\)\textsubscript{consensus}\\
\hline
HINT HT & 13\,619 & 142\,541 & 16\,706 & 0.46 & 0.066\\
\hline
Consensus & 55 & 117 & 74\,062 & 0.0051 & 1\\
dmGWAS & 194 & 450 & 49\,115 & 0.19 & 0.41\\
heinz & 4 & 3 & 113\,633 & 0.0012 & 0.21\\
HotNet2 & 440 & 374 & 7\,739 & 0.048 & 0.31\\
LEAN & 0 & 0 & - & - & -\\
SConES GI & 0 (1) & 0 & - & - & -\\
SigMod & 142 & 249 & 92\,603 & 0.0083 & 0.73\\
\end{tabular}
\begin{tablenotes}
  \footnotesize{
    \item \textbf{\emph{\# genes}}: number of genes selected out of those that are part of the PPIN; for SConES GI the total number of genes, including RNA genes, was added in parentheses. \textbf{\mean{Betweenness}}: mean betweenness of the selected genes in the PPIN. \textbf{\median{P}\textsubscript{gene}}: median P-value of the selected genes. \textbf{\(\rho\)\textsubscript{consensus}}: Pearson correlation between the subnetwork and the consensus network.
  }
\end{tablenotes}
\end{threeparttable}
\end{table}
